Live and Let Die? (Delayed Issues not Harder to Resolve)

### 2016-09-16 data.tex

One reason to use the SEI data for the analysis of this paper is its remarkably low level of noise.
Nichols et al.~\cite{shirai14}  report that
the noise levels in the SEI TSP data are smaller than those seen
in other data sets. They found in the SEI TSP data that:\bi 
\item
4\% of the data was incorrect (e.g. nulls, illegal formats);
\item  2\% of the data has inconsistencies such as timestamps
where the stop time was before the start time;
\item 3\% of the data contained values that were not credible
such as tasks listed in one day that took more than six hours for a single developer.
\ei 
One explanation for this low level of noise is the TSP process.
One the guiding principles of TSP was that  people performing the work are  responsible for planning and tracking the work. That is,  all the data collected here was entered
by local developers. This data was then check by local {\em coaches} before being sent to the SEI
databases. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data.
The use of certified local coaches within each project increases the integrity of our data.

@@

One special feature of  \fig{waterfall} is
the {\em before} where the   TSP Launch Preparation scripts and checklists (see TSP: Leading a development Team~\cite{Humphrey:2005}) require the TSP team to assure that management and business have clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model. For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.

@@

Note that, in that Figure~\ref{fig:waterfall},
several  phases in which the product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
In TSP  reviews,  individuals perform personal reviews of their work products prior to the peer review
(which TSP calls the inspection(.
Also, Figure~\ref{fig:waterfall} divides  testing     as follows. Developers perform unit test prior to code complete.  After code complete a standard phase is the integration, which combines program units into workable system ready for system test. Integration,  system test, and acceptance test are often performed by another group.

>>>>

One reason to use the SEI data for the analysis of this paper is its remarkably low level of noise.
Nichols et al.~\cite{shirai14}  report that
the noise levels in the SEI TSP data are smaller than those seen
in other data sets. They found in the SEI TSP data that:\bi 
\item
4\% of the data was incorrect (e.g. nulls, illegal formats);
\item  2\% of the data has inconsistencies such as timestamps
where the stop time was before the start time;
\item 3\% of the data contained values that were not credible
such as tasks listed in one day that took more than six hours for a single developer.
\ei 
One explanation for this low level of noise is the TSP process.
One the guiding principles of TSP was that  people performing the work are  responsible for planning and tracking the work. That is,  all the data collected here was entered
by local developers. This data was then checked by local coaches before being sent to the SEI
databases. Coaches are certified by demonstrating competent use of the TSP process with the artifacts and data.
The use of certified local coaches within each project increases the integrity of our data.

@@

One special feature of  \fig{waterfall} is
the {\em before} phase, in which the TSP team assures that management has clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model ~\cite{Humphrey:2005}. For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.

@@

Note that, in that Figure~\ref{fig:waterfall},
several  phases in which the product is created have sub-phases of {\em review} and {\em inspect} to remove defects.
In TSP  reviews,  individuals perform personal reviews of their work products prior to the peer review
(which TSP calls the inspection).
Also, Figure~\ref{fig:waterfall} divides  testing     as follows. Developers perform unit test prior to code complete.  After code complete a standard phase is integration, which combines program units into a workable system ready for system test. Integration,  system test, and acceptance test are often performed by another group.

### 2016-09-16 phaseDelay.tex

\section{Introduction}
In 2013-2014, 
eleven  million programmers~\cite{pettey14} and
half a trillion dollars~\cite{avram14} were spent on information technology.
Such a large and growing effort should be managed and optimized via  well-researched conclusions.

It is standard practice
in other fields, such as medicine,
to continually revisit old conclusions~\cite{prasad13}.
Accordingly, this paper revisits
the widely-held belief of a {\em delayed issue effect} (hereafter, DIE).
Later in this paper, we offer a precise definition for this effect.
For the moment, we describe it as follows:
it is far easier to resolve issues earlier rather than later in the lifecycle.
 \fig{b81} shows an example of the delayed issue effect (relating
 the relative cost of fixing requirements issues at different phases of a project).

 Basili \& Boehm comment that since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[angle=270,width=3in]{img/b81.pdf}
 \end{center}
 \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}
Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.  McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a proven
fact.
For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

@@

To be clear, we do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold. Our explanation of the observed lack-of-effect is five-fold:
\be
\item The effect might be an historical relic. Evidence:
the effect was first described in the era of punch card computing and non-interactive environments.
\item The effect might have been intermittent (rather than some fundamental law of software). Evidence: we can  found nearly
as many papers reporting the effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76} as otherwise~\cite{Royce98,Boehm80,Shull02}.
\item The effect might be confined to very large systems- in which case it would be
acceptable during development to let smaller to medium
sized projects carry some unresolved issues from early phases into later phases.
\item The effect might be mitigated by modern software development approaches that
encourage change and revision of older parts of the system.
\item The effect might be mitigated by modern software development tools
that simplify the process of large-scale reorganization of software systems.
\ee
 %One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method. 
Our results beg the question: why did an idea, with so little support, become so widespread in the software engineering literature? No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years. Possibly the concept of the delayed issue effect (or its more specific description: requirements errors are the hardest to fix)
has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}. 
Nevertheless, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified. 
Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.

>>>>

\section{Introduction}
In 2013-2014, 
eleven  million programmers~\cite{pettey14} and
half a trillion dollars~\cite{avram14} were spent on information technology.
Such a large and growing effort should be managed and optimized via  well-researched conclusions.  
To assist in achieving this, there has been a growing recognition within the software engineering research community 
of the importance of theory building [Sjoberg08, Paivarinta15, Stol15]. A good theory allows empirical research to go beyond simply
reporting observations and instead provides explanations for why results are observed [Stol15]. Theory-building needs to be an iterative process,
in which results from practice are used to refine theories and theories are used to inform future observation and data collection [Paivarinta15, Stol15]. 
It is no coincidence that it  
is standard practice
in other fields, such as medicine,
to continually revisit old conclusions in the light of new theories and new understanding~\cite{prasad13}.

Accordingly, this paper revisits
a fundamental software engineering theory, the {\em delayed issue effect} (hereafter, DIE).
Later in this paper, we offer a precise definition for this effect.
For the moment, we describe it as follows:
it is far easier to resolve issues earlier rather than later in the lifecycle.
 \fig{b81} shows an example of the delayed issue effect (relating
 the relative cost of fixing requirements issues at different phases of a project).

This theory has been used as the basis for decision-making in software engineering. For example, Basili \& Boehm comment that since the 1980s,  this effect
 \begin{quote}
``...has been a major driver in focusing
industrial software practice on thorough
requirements analysis and design,
on early verification and validation, and
on up-front prototyping and simulation
to avoid costly downstream fixes''~\cite{boehm01}.
\end{quote}
 \begin{figure}[!b]  
\begin{center}
\includegraphics[angle=270,width=3in]{img/b81.pdf}
 \end{center}
 \caption{An example DIE effect. From Boehm'81~\cite{Boehm81}. }\label{fig:b81}
 \end{figure}
Other prominent authors have commented on its perceived usefulness as a rule of thumb for software engineers.  McConnell mentions it as a "common observation" in the field and  summarizes the intuitive argument for why it should be so: 
\begin{quote}
''A small mistake in upstream work can affect large amounts of downstream work. A change to a single sentence in a requirements specification can imply changes in hundreds of lines of code spread across numerous classes or modules, dozens of test cases, and numerous pages of end-user documentation''~\cite{mcconnell01}. 
\end{quote}
Glass also endorses this effect, asserting that ``requirements errors are the most expensive to fix when found during production but the cheapest to fix early in development'' is ``really just common sense''~\cite{glass02}.  Other researchers
are just as adamant in asserting that that the delayed issue effect is a proven
fact.
For example, what we call the delayed issued effect was listed first by Boehm and Basili in their "Top 10 list" of ``objective and quantitative data, relationships,
and predictive models that help
software developers avoid predictable pitfalls
and improve their ability to predict
and control efficient software projects''~\cite{boehm01}.

@@

To be clear, we do not claim that this effect {\em never} holds in software projects; just that it cannot be assumed to {\em always} hold. Our explanation of the observed lack-of-effect is five-fold:
\be
\item The effect might be an historical relic. Evidence:
the effect was first described in the era of punch card computing and non-interactive environments.
\item The effect might have been intermittent (rather than some fundamental law of software). Evidence: we can  found nearly
as many papers reporting the effect~\cite{Boehm76,Boehm81,steck04,Fagan76,Stephenson76} as otherwise~\cite{Royce98,Boehm80,Shull02}.
\item The effect might be confined to very large systems- in which case it would be
acceptable during development to let smaller to medium
sized projects carry some unresolved issues from early phases into later phases.
\item The effect might be mitigated by modern software development approaches that
encourage change and revision of older parts of the system.
\item The effect might be mitigated by modern software development tools
that simplify the process of large-scale reorganization of software systems.
\ee
 %One hypothesized explanation that can be ruled out is that the lack of a phase delay effect is due to using the TSP method. 
Our results raise the question: why did an idea, with so little support, become so widespread in the software engineering literature? No doubt the original evidence was compelling at the time, but much has changed in the realm of software development in the subsequent 40 years. Possibly the concept of the delayed issue effect (or its more specific description: requirements errors are the hardest to fix)
has persisted because, to use Glass's terms on the subject, it seems to be ``just common sense''\cite{glass02}. 
Nevertheless, in a rapidly changing field such as software engineering, even widely-held rules of thumb must be periodically re-verified. 
Progress in the domain of software analytics has made such periodic checks more cost-effective and feasible, and we argue that an examination of local behaviors (rather than simply accepting global heuristics) can be of significant benefit.

### 2016-09-16 tex/responses.tex

\begin{quote} {\em Thank you for the careful reviews. Using
those comments we have extensively revised and extended this draft.
Specifically, the old version was 20 pages and this version is 27 pages (and the new
material all comes from the clarification and correction points offered by the reviewers).
We hope that  you will find this draft greatly improved.}
\end{quote}

@@

\begin{quote}{\em This new draft now maps our work into the SE literature on theory building. This new material appears in the introduction as well as being
peppered throughout the document (in numerous places).}\end{quote}

@@

\begin{quote}{\em We now map our work to prior publications on theory-building in software engineering. That mapping appears in numerous comments through out the text (using the references offered by this reviewer) with the bulk of the comments being top of p2 and the conclusion.

@@

\begin{quote}{\em We agree with the reviewer - the phrasing of our claims and the supporting evidence were not congruent. Our intention is to argue from multiple angles that the evidence for DIE is sparse and, more importantly, that DIE persists as a popular belief. Our goal here is to motivate that we should re-investigate the DIE. We have toned down the language and combined our arguments on ``widely-believed'' and ``poorly-documented'' into a new \tion{belief} whose focus is to motivate our study.}\end{quote}

@@

\begin{quote}{\em An astute observation. We provide more information on the types of projects included in our sample in \tion{data_character}

@@

\begin{quote}{\em We understand why you are making that suggestion.
That figure is something of a classic. It is appropriate we chop it like that?}.
\end{quote}

@@

\begin{quote}{\em
Thanks for those pointers. We will fix.
 }\end{quote}

@@

A very minor issue: why is Figure 2 [NOTE: now \fig{survey_results}] a ... figure, instead of
a table? Same for all other tables ...
\begin{quote}
{\em Authors' preference. When a document has many tables and figures, navigating
between them is hard. Calling them all "figures" makes navigation faster. 
But do you want us to change that?}
\end{quote}
Section 5.1

>>>>

\begin{quote} {\em Thank you for the careful reviews. Using
those comments we have extensively revised and extended this draft.
Specifically, the old version was 20 pages and this version is 29 pages (and the new
material all comes from the clarification and correction points offered by the reviewers).
We hope that  you will find this draft greatly improved.}
\end{quote}

@@

\begin{quote}{\em This new draft now maps our work into the SE literature on theory building. This new material appears in the introduction as well as throughout the document.}\end{quote}

@@

\begin{quote}{\em We now map our work to prior publications on theory-building in software engineering, as well as theory falsification in general. That mapping appears in numerous comments through out the text (using the references offered by this reviewer) with the bulk of the comments being top of p2 and the conclusion.

@@

\begin{quote}{\em We agree with the reviewer - the phrasing of our claims and the supporting evidence were not congruent. Our intention is to argue from multiple angles that the evidence for DIE is sparse and, more importantly, that DIE persists as a popular belief. Our goal here is to motivate that we should re-investigate the DIE. We have toned down the language and combined our arguments on ``widely-believed'' and ``poorly-documented'' into a new \tion{belief} whose focus is to motivate our study. We have added more details on our literature search and snowball sampling. We have also highlighted how blog articles often cite secondary sources, which in turn cite Software Engineering Economics, to illustrate how the DIE pehnemonon has perpetuated.}\end{quote}

@@

\begin{quote}{\em We provide more information on the types of projects included in our sample in \tion{data_character}

@@

\begin{quote}{\em We understand why you are making that suggestion.
That figure is something of a classic. Therefore we opted not to chop the figure but hope the text is sufficiently clear}.
\end{quote}

@@

\begin{quote}{\em
Thanks for those pointers. We have fixed the issue.
 }\end{quote}

@@

A very minor issue: why is Figure 2 [NOTE: now \fig{survey_results}] a ... figure, instead of
a table? Same for all other tables ...
\begin{quote}
{\em Authors' preference. When a document has many tables and figures, navigating
between them is hard. Calling them all "figures" makes navigation faster. }
\end{quote}
Section 5.1
